Parameter Recovery Validation

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): use T_M instead of max
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

\begin{align}
T_{\max} &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \quad \text{(process duration)} \\
t^*_j &\sim \text{Uniform}(0, T_{\max}) \quad \text{(cell temporal positions)} \\
t^*_{0,\text{on}i} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \quad \text{(absolute activation onset time)} \\
\delta^*_i &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \quad \text{(absolute activation duration)}
\end{align}

@@

- $t^*_{0,\text{on}i} < 0.0$ (absolute activation onset before observation window)
- $R_{\text{on},i} > 2.0$ (moderate to strong fold-change during activation)

@@

- $0 < t^*_{0,\text{on}i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
- $\delta^*_i < 2.0$ (short activation duration relative to process, assuming conservative prior Gamma(5,1))
- $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)

@@

- $0 < t^*_{0,\text{on}i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
- $\delta^*_i > 2.5$ (long activation duration relative to process, assuming conservative prior Gamma(5,1))
- $R_{\text{on},i} > 2.0$ (strong fold-change during activation)

@@

The process duration $T_{\max}$ determines the temporal scope of observable dynamics and requires careful hyperparameter selection based on biological expectations and experimental context.

@@

In the dimensionless framework where the reference rate is 1, the process duration $T_{\max}$ determines what fraction of steady-state dynamics will be observable:

$$\text{Steady-state approach} = 1 - e^{-T_{\max}}$$

@@

- $T_{\max} = 1$: Observes 63% approach to steady state
- $T_{\max} = 3$: Observes 95% approach to steady state
- $T_{\max} = 5$: Observes >99% approach to steady state

@@

**Conservative Prior (Expect Complete Dynamics):**
$$T_{\max} \sim \text{Gamma}(5.0, 1.0) \quad \text{(mean = 5, variance = 5)}$$

@@

**Permissive Prior (Let Data Determine Range):**
$$T_{\max} \sim \text{Gamma}(1.0, 0.2) \quad \text{(mean = 5, variance = 25)}$$

@@

**Skeptical Prior (Assume Short Processes):**
$$T_{\max} \sim \text{Gamma}(1.0, 1.0) \quad \text{(mean = 1, variance = 1, heavy tail)}$$

@@

With independent parameterization, gene timing parameters should be calibrated to $T_{\max}$ expectations:

\begin{align}
t^*_{0,\text{on}i} &\sim \text{Normal}(0.3 \mathbb{E}[T_{\max}], \sigma_{on}^2) \quad \text{(expect activation at 30% of process)} \\
\delta^*_i &\sim \text{LogNormal}(\log(0.4 \mathbb{E}[T_{\max}]), \sigma_{\delta}^2) \quad \text{(expect 40% duration activation)}
\end{align}

@@

1. **Gene parameters are sampled independently**: $t^*_{0,\text{on}i} \not\propto T_{\max}$ and $\delta^*_i \not\propto T_{\max}$

2. **Scaling cell times but not gene parameters changes piecewise phase assignments**:
   - Original: Cell $j$ in phase determined by comparing $t^*_j$ vs. $t^*_{0,\text{on}i}$
   - Scaled: Cell $j$ in phase determined by comparing $\lambda t^*_j$ vs. $t^*_{0,\text{on}i}$ (unchanged)

3. **The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{0,\text{on}i}$**:
   - Phase boundaries change when cell times are scaled but gene times are not
   - Different phase assignments lead to different analytical solution forms

4. **Therefore**: $\mathcal{L}(T_{\max}, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_{\max}, \{\lambda t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\})$ □

@@

1. **Phase determination remains consistent**: For any observed time $t^* \geq 0$, we can still determine which phase applies by comparing $t^*$ to $t^*_{0,\text{on}i}$ and $t^*_{0,\text{on}i} + \delta^*_i$

@@

- **Phase 1 may be unobserved**: If $t^*_{0,\text{on}i} < 0$, no observed cells will be in Phase 1
- **Phase 2 starts immediately**: Cells at $t^* = 0$ may already be in the activated state
- **Duration constraints**: For pre-activation genes, $\delta^*_i > |t^*_{0,\text{on}i}|$ ensures activation extends into the observation window

@@

For **Normal parameters** ($t^*_{0,\text{on}i}$), the probability of satisfying constraint $X > \theta$ is:

@@

# Prior hyperparameters (updated for independent absolute parameterization)
# Assuming T_max ~ Gamma(5.0, 1.0) with E[T_max] = 5.0
E_T_max = 5.0
priors = {
    'R_on': {'loc': 0.916, 'scale': 0.4},                           # log(2.5), fold-change (LogNormal)
    't_on_star': {'loc': 0.3 * E_T_max, 'scale': 0.8},             # Normal(1.5, 0.8²), absolute onset
    'delta_star': {'loc': np.log(0.4 * E_T_max), 'scale': 0.45},   # log(2.0), absolute duration (LogNormal)
}

# Calculate constraint feasibility for each pattern (updated for independent parameterization)
patterns = {
    'pre_activation': [
        ('t_on_star', '<', 0.0),        # Activation before observation window
        ('R_on', '>', 2.0),             # Moderate to strong fold change
    ],
    'transient': [
        ('t_on_star', '>', 0.0),                    # Activation within observation window
        ('t_on_star', '<', 0.3 * E_T_max),         # Early enough to complete cycle
        ('delta_star', '<', 0.4 * E_T_max),        # Short pulse duration
        ('R_on', '>', 2.0),                        # Sufficient fold change to observe
    ],
    'sustained': [
        ('t_on_star', '>', 0.0),                    # Activation within observation window
        ('t_on_star', '<', 0.3 * E_T_max),         # Early activation onset
        ('delta_star', '>', 0.5 * E_T_max),        # Long activation duration
        ('R_on', '>', 2.0),                        # Strong fold change
    ]
}

@@

# independent time parameters
pgm.add_node("T_max", r"$T_{\max}$", 5.5, 5)

@@

# independent time structure edges
pgm.add_edge("alpha_T", "T_max", plot_params=edge_params)
pgm.add_edge("beta_T", "T_max", plot_params=edge_params)

# independent time edges (no hierarchical scaling)
pgm.add_edge("T_max", "t_star_j", plot_params=edge_params)

@@

- $\theta_i$ represents the vector of **non-temporal** piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$
- $\Delta_i$ represents the vector of **independent temporal** gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$
- The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
- The hyperparameters $\mu_{\Delta}, \sigma_{\Delta}^2$ represent the collection of hyperparameters for all components of $\Delta_i$
- $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable
- $T_{\max}$ determines the process duration and constrains cell temporal positions via $t^*_j \sim \text{Uniform}(0, T_{\max})$

@@

**Independent temporal parameters:**
\begin{align}
    T_{\max} &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \quad \text{(process duration)} \\
    t^*_j &\sim \text{Uniform}(0, T_{\max}) \quad \text{(cell temporal positions)} \\
    t^*_{0,\text{on}i} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \quad \text{(absolute activation onset time)} \\
    \delta^*_i &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \quad \text{(absolute activation duration)}
\end{align}

@@

- **Pre-activation**: $t^*_{0,\text{on}i} < 0.0$ and $R_{\text{on},i} > 2.0$
- **Transient**: $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$, $\delta^*_i < 0.4 \mathbb{E}[T_{\max}]$, and $R_{\text{on},i} > 2.0$
- **Sustained**: $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$, $\delta^*_i > 0.5 \mathbb{E}[T_{\max}]$, and $R_{\text{on},i} > 2.0$

@@

- **$T_{\max} \sim \text{Gamma}(\alpha_{T}, \beta_{T})$**: Process duration with hyperparameters chosen based on experimental context (conservative: Gamma(5,1), permissive: Gamma(1,0.2), skeptical: Gamma(1,1)).

- **$t^*_{0,\text{on}i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around 1.5 to expect activation at 30% of process duration (assuming conservative prior Gamma(5,1) with mean 5). **Negative values** enable pre-activation scenarios where genes were already activated before observation began.

@@

- **Pre-activation** ($t^*_{0,\text{on}i} < 0$): Probability depends on $\mu_{on}$ and $\sigma_{on}$ choices
- **Transient** ($0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i < 2.0$): Balanced probability for observable complete cycles (assuming conservative prior Gamma(5,1))
- **Sustained** ($0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i > 2.5$): Balanced probability for long-duration activation (assuming conservative prior Gamma(5,1))

@@

The dimensionless concentrations are computed using the analytical solutions for each phase as derived above, with the appropriate phase determined by comparing $t^*_j$ to the gene-specific activation parameters $t^*_{0,\text{on}i}$ and $\delta^*_i$. All temporal parameters are now independent, eliminating scaling symmetry.

@@

where $\theta$ represents the complete parameter vector including all gene-specific parameters $(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i})$, independent time parameters $(T_{\max}, t^*_j)$, and cell-specific capture efficiencies $(\lambda_j)$.

@@

\begin{align}
p(\tilde{y}) &= \int p(\tilde{y} | \gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i}, T_{\max}, t^*_j, \lambda_j) \times \nonumber \\
&\quad p(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i}, T_{\max}, t^*_j, \lambda_j) \, d\theta \label{eq-prior-predictive-piecewise}
\end{align}

@@

\begin{align}
p(\tilde{y}) &= \int \int \int \int p(\tilde{y} | T_{\max}, \theta_{\text{gene}}, \theta_{\text{cell}}, \theta_{\text{tech}}) \times \nonumber \\
&\quad p(T_{\max}) \, p(\theta_{\text{gene}}) \, p(\theta_{\text{cell}}) \, p(\theta_{\text{tech}}) \, dT_{\max} \, d\theta_{\text{gene}} \, d\theta_{\text{cell}} \, d\theta_{\text{tech}} \label{eq-independent-integration}
\end{align}

@@

- $\theta_{\text{gene}} = \{\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i}\}_{i=1}^G$ represents gene-specific parameters
- $\theta_{\text{cell}} = \{t^*_j\}_{j=1}^N$ represents cell-specific temporal parameters
- $\theta_{\text{tech}} = \{\lambda_j\}_{j=1}^N$ represents technical efficiency parameters

@@

1. **Sample process duration**:
   \begin{align}
   {T_{\max}}^{(s)} &\sim \text{Gamma}(\alpha_{T}, \beta_{T})
   \end{align}

2. **Sample cell-specific temporal positions**:
   \begin{align}
   {t^{*}_j}^{(s)} &\sim \text{Uniform}(0, {T_{\max}}^{(s)}) \quad \text{for } j \in \{1, \ldots, N\}
   \end{align}

3. **Sample gene-specific parameters**:
   \begin{align}
   {\gamma^{*}_i}^{(s)} &\sim \text{LogNormal}(\log(1.0), 0.5^2) \\
   {R_{\text{on},i}}^{(s)} &\sim \text{LogNormal}(\log(2.5), 0.4^2) \\
   {t^*_{0,\text{on}i}}^{(s)} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \\
   {\delta^*_i}^{(s)} &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \\
   {U_{0i}}^{(s)} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)
   \end{align}

@@

5. **Compute latent concentrations**:
   \begin{align}
   {u^{*}_{ij}}^{(s)}, {s^{*}_{ij}}^{(s)} &= f_{\text{piecewise}}({t^*_j}^{(s)}, {\gamma^*_i}^{(s)}, {R_{\text{on},i}}^{(s)}, {t^*_{0,\text{on}i}}^{(s)}, {\delta^*_i}^{(s)})
   \end{align}

@@

1. **Process duration uncertainty**: Uncertainty in $T_{\max}$ affects the temporal scope of the biological process and constrains cell temporal positions

@@

4. **Gene-specific temporal pattern uncertainty**: Prior uncertainty in $t^*_{0,\text{on}i}, \delta^*_i$ affects when activation occurs and how long it lasts in absolute time

@@

1. **Parameter uncertainty**: $\text{Var}[\theta | y_{obs}]$ from the posterior distribution over all model parameters
2. **Temporal uncertainty**: Uncertainty in latent time coordinates $t^*_j$ and process duration $T_{\max}$
3. **Kinetic uncertainty**: Uncertainty in gene-specific kinetic parameters $\gamma^*_i$ and fold-changes $R_{\text{on},i}$
4. **Concentration uncertainty**: Uncertainty in predicted latent concentrations $u^*_{ij}(\theta), s^*_{ij}(\theta)$ arising from parameter uncertainty
5. **Observation uncertainty**: Poisson noise in final count observations

@@

- **Non-temporal gene parameters**: $\gamma^*_i \sim \text{LogNormal}(\log(1.0), 0.5^2)$, $R_{\text{on},i} \sim \text{LogNormal}(\log(2.5), 0.4^2)$
- **Independent temporal gene parameters**: $t^*_{0,\text{on}i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$, $\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$
- **Characteristic scales**: $U_{0i} \sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)$
- **Process duration**: $T_{\max} \sim \text{Gamma}(\alpha_{T}, \beta_{T})$
- **Cell-specific times**: $t^*_j \sim \text{Uniform}(0, T_{\max})$
- **Capture efficiencies**: $\lambda_j \sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)$

@@

$$u^*_{ij}(\theta), s^*_{ij}(\theta) = f_{\text{piecewise}}(t^*_j, \gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i)$$

@@

1. Compute phase assignments for each cell-gene pair using ${{t^{\ast}_{j}}}^{(s)}$, ${{t^{\ast}_{0,\text{on}i}}}^{(s)}$, ${{\delta^{\ast}_i}}^{(s)}$
2. Evaluate analytical solutions: ${{u^*_{ij}}}^{(s)}, {{s^*_{ij}}}^{(s)} = f_{\text{piecewise}}({{t^*_{j}}}^{(s)}, {{\gamma^*_i}}^{(s)}, {{R_{\text{on},i}}}^{(s)}, {{t^*_{0,\text{on}i}}}^{(s)}, {{\delta^*_i}}^{(s)})$
3. Sample observations:

@@

1. **Kinetic parameter uncertainty**: Posterior distributions over $\gamma^*_i$ and $R_{\text{on},i}$ affect the shape and magnitude of concentration trajectories
2. **Temporal parameter uncertainty**: Uncertainty in $t^*_{0,\text{on}i}$ and $\delta^*_i$ affects when activation occurs and how long it lasts in absolute time
3. **Process duration uncertainty**: Uncertainty in $T_{\max}$ affects the temporal scope and constrains cell temporal positions
4. **Cell-specific uncertainty**: Uncertainty in individual $t^*_j$ and $\lambda_j$ affects cell-specific predictions
5. **Scale uncertainty**: Uncertainty in $U_{0i}$ affects the absolute magnitude of predicted counts

@@

1. **Recover independent piecewise parameters**: $\gamma^*_i$, $R_{\text{on},i}$, $t^*_{0,\text{on}i}$, $\delta^*_i$
2. **Maintain process duration inference**: Recover $T_{\max}$ and cell temporal positions $t^*_j$
3. **Handle realistic noise levels**: Maintain accuracy with Poisson observation noise
4. **Scale efficiently**: Demonstrate computational advantages of eliminated scaling symmetry
5. **Validate fold-change interpretation**: Ensure $R_{\text{on},i}$ estimates match true activation strength
6. **Assess dimensional consistency**: Verify elimination of parameter redundancy through fixed $\alpha^*_{\text{off}i} = 1.0$
7. **Validate parameter identifiability**: Confirm elimination of scaling symmetry through independent parameterization
8. **Assess parameter separation**: Verify clean separation between kinetic ($\gamma^*_i$, $R_{\text{on},i}$) and temporal ($t^*_{0,\text{on}i}$, $\delta^*_i$) parameter inference
9. **Validate independent parameter framework**: Confirm that independent temporal parameters enable reliable parameter recovery across different experimental contexts

>>>>

\begin{align}
T_M &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \quad \text{(process duration)} \\
t^*_j &\sim \text{Uniform}(0, T_M) \quad \text{(cell temporal positions)} \\
t^*_{on,i} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \quad \text{(absolute activation onset time)} \\
\delta^*_i &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \quad \text{(absolute activation duration)}
\end{align}

@@

- $t^*_{on,i} < 0.0$ (absolute activation onset before observation window)
- $R_{\text{on},i} > 2.0$ (moderate to strong fold-change during activation)

@@

- $0 < t^*_{on,i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
- $\delta^*_i < 2.0$ (short activation duration relative to process, assuming conservative prior Gamma(5,1))
- $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)

@@

- $0 < t^*_{on,i} < 1.5$ (early activation within process duration, assuming conservative prior Gamma(5,1))
- $\delta^*_i > 2.5$ (long activation duration relative to process, assuming conservative prior Gamma(5,1))
- $R_{\text{on},i} > 2.0$ (strong fold-change during activation)

@@

The process duration $T_M$ determines the temporal scope of observable dynamics and requires careful hyperparameter selection based on biological expectations and experimental context.

@@

In the dimensionless framework where the reference rate is 1, the process duration $T_M$ determines what fraction of steady-state dynamics will be observable:

$$\text{Steady-state approach} = 1 - e^{-T_M}$$

@@

- $T_M = 1$: Observes 63% approach to steady state
- $T_M = 3$: Observes 95% approach to steady state
- $T_M = 5$: Observes >99% approach to steady state

@@

**Conservative Prior (Expect Complete Dynamics):**
$$T_M \sim \text{Gamma}(5.0, 1.0) \quad \text{(mean = 5, variance = 5)}$$

@@

**Permissive Prior (Let Data Determine Range):**
$$T_M \sim \text{Gamma}(1.0, 0.2) \quad \text{(mean = 5, variance = 25)}$$

@@

**Skeptical Prior (Assume Short Processes):**
$$T_M \sim \text{Gamma}(1.0, 1.0) \quad \text{(mean = 1, variance = 1, heavy tail)}$$

@@

With independent parameterization, gene timing parameters should be calibrated to $T_M$ expectations:

\begin{align}
t^*_{on,i} &\sim \text{Normal}(0.3 \times 5, \sigma_{on}^2) = \text{Normal}(1.5, \sigma_{on}^2) \quad \text{(expect activation at 30% of process)} \\
\delta^*_i &\sim \text{LogNormal}(\log(0.4 \times 5), \sigma_{\delta}^2) = \text{LogNormal}(\log(2.0), \sigma_{\delta}^2) \quad \text{(expect 40% duration activation)}
\end{align}

@@

1. **Gene parameters are sampled independently**: $t^*_{on,i} \not\propto T_M$ and $\delta^*_i \not\propto T_M$

2. **Scaling cell times but not gene parameters changes piecewise phase assignments**:
   - Original: Cell $j$ in phase determined by comparing $t^*_j$ vs. $t^*_{on,i}$
   - Scaled: Cell $j$ in phase determined by comparing $\lambda t^*_j$ vs. $t^*_{on,i}$ (unchanged)

3. **The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{on,i}$**:
   - Phase boundaries change when cell times are scaled but gene times are not
   - Different phase assignments lead to different analytical solution forms

4. **Therefore**: $\mathcal{L}(T_M, \{t^*_j\}, \{t^*_{on,i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_M, \{\lambda t^*_j\}, \{t^*_{on,i}\}, \{\delta^*_i\})$ □

@@

When $t^*_{on,i} < 0$:

1. **Phase determination remains consistent**: For any observed time $t^* \geq 0$, we can still determine which phase applies by comparing $t^*$ to $t^*_{on,i}$ and $t^*_{on,i} + \delta^*_i$

@@

- **Phase 1 may be unobserved**: If $t^*_{on,i} < 0$, no observed cells will be in Phase 1
- **Phase 2 starts immediately**: Cells at $t^* = 0$ may already be in the activated state
- **Duration constraints**: For pre-activation genes, $\delta^*_i > |t^*_{on,i}|$ ensures activation extends into the observation window

@@

For **Normal parameters** ($t^*_{on,i}$), the probability of satisfying constraint $X > \theta$ is:

@@

# Prior hyperparameters (updated for independent absolute parameterization)
# Assuming T_M ~ Gamma(5.0, 1.0) with E[T_M] = 5.0
E_T_M = 5.0
priors = {
    'R_on': {'loc': 0.916, 'scale': 0.4},                           # log(2.5), fold-change (LogNormal)
    't_on_star': {'loc': 1.5, 'scale': 0.8},                       # Normal(1.5, 0.8²), absolute onset
    'delta_star': {'loc': np.log(2.0), 'scale': 0.45},             # log(2.0), absolute duration (LogNormal)
}

# Calculate constraint feasibility for each pattern (updated for independent parameterization)
patterns = {
    'pre_activation': [
        ('t_on_star', '<', 0.0),        # Activation before observation window
        ('R_on', '>', 2.0),             # Moderate to strong fold change
    ],
    'transient': [
        ('t_on_star', '>', 0.0),                    # Activation within observation window
        ('t_on_star', '<', 1.5),                   # Early enough to complete cycle
        ('delta_star', '<', 2.0),                  # Short pulse duration
        ('R_on', '>', 2.0),                        # Sufficient fold change to observe
    ],
    'sustained': [
        ('t_on_star', '>', 0.0),                    # Activation within observation window
        ('t_on_star', '<', 1.5),                   # Early activation onset
        ('delta_star', '>', 2.5),                  # Long activation duration
        ('R_on', '>', 2.0),                        # Strong fold change
    ]
}

@@

# independent time parameters
pgm.add_node("T_M", r"$T_M$", 5.5, 5)

@@

# independent time structure edges
pgm.add_edge("alpha_T", "T_M", plot_params=edge_params)
pgm.add_edge("beta_T", "T_M", plot_params=edge_params)

# independent time edges (no hierarchical scaling)
pgm.add_edge("T_M", "t_star_j", plot_params=edge_params)

@@

- $\theta_i$ represents the vector of **non-temporal** piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$
- $\Delta_i$ represents the vector of **independent temporal** gene parameters $(t^*_{on,i}, \delta^*_i)$
- The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
- The hyperparameters $\mu_{\Delta}, \sigma_{\Delta}^2$ represent the collection of hyperparameters for all components of $\Delta_i$
- $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable
- $T_M$ determines the process duration and constrains cell temporal positions via $t^*_j \sim \text{Uniform}(0, T_M)$

@@

**Independent temporal parameters:**
\begin{align}
    T_M &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \quad \text{(process duration)} \\
    t^*_j &\sim \text{Uniform}(0, T_M) \quad \text{(cell temporal positions)} \\
    t^*_{on,i} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \quad \text{(absolute activation onset time)} \\
    \delta^*_i &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \quad \text{(absolute activation duration)}
\end{align}

@@

- **Pre-activation**: $t^*_{on,i} < 0.0$ and $R_{\text{on},i} > 2.0$
- **Transient**: $0 < t^*_{on,i} < 1.5$, $\delta^*_i < 2.0$, and $R_{\text{on},i} > 2.0$ (assuming conservative prior Gamma(5,1))
- **Sustained**: $0 < t^*_{on,i} < 1.5$, $\delta^*_i > 2.5$, and $R_{\text{on},i} > 2.0$ (assuming conservative prior Gamma(5,1))

@@

- **$T_M \sim \text{Gamma}(\alpha_{T}, \beta_{T})$**: Process duration with hyperparameters chosen based on experimental context (conservative: Gamma(5,1), permissive: Gamma(1,0.2), skeptical: Gamma(1,1)).

- **$t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around 1.5 to expect activation at 30% of process duration (assuming conservative prior Gamma(5,1) with mean 5). **Negative values** enable pre-activation scenarios where genes were already activated before observation began.

@@

- **Pre-activation** ($t^*_{on,i} < 0$): Probability depends on $\mu_{on}$ and $\sigma_{on}$ choices
- **Transient** ($0 < t^*_{on,i} < 1.5$ and $\delta^*_i < 2.0$): Balanced probability for observable complete cycles (assuming conservative prior Gamma(5,1))
- **Sustained** ($0 < t^*_{on,i} < 1.5$ and $\delta^*_i > 2.5$): Balanced probability for long-duration activation (assuming conservative prior Gamma(5,1))

@@

The dimensionless concentrations are computed using the analytical solutions for each phase as derived above, with the appropriate phase determined by comparing $t^*_j$ to the gene-specific activation parameters $t^*_{on,i}$ and $\delta^*_i$. All temporal parameters are now independent, eliminating scaling symmetry.

@@

where $\theta$ represents the complete parameter vector including all gene-specific parameters $(\gamma^*_i, R_{\text{on},i}, t^*_{on,i}, \delta^*_i, U_{0i})$, independent time parameters $(T_M, t^*_j)$, and cell-specific capture efficiencies $(\lambda_j)$.

@@

\begin{align}
p(\tilde{y}) &= \int p(\tilde{y} | \gamma^*_i, R_{\text{on},i}, t^*_{on,i}, \delta^*_i, U_{0i}, T_M, t^*_j, \lambda_j) \times \nonumber \\
&\quad p(\gamma^*_i, R_{\text{on},i}, t^*_{on,i}, \delta^*_i, U_{0i}, T_M, t^*_j, \lambda_j) \, d\theta \label{eq-prior-predictive-piecewise}
\end{align}

@@

\begin{align}
p(\tilde{y}) &= \int \int \int \int p(\tilde{y} | T_M, \theta_{\text{gene}}, \theta_{\text{cell}}, \theta_{\text{tech}}) \times \nonumber \\
&\quad p(T_M) \, p(\theta_{\text{gene}}) \, p(\theta_{\text{cell}}) \, p(\theta_{\text{tech}}) \, dT_M \, d\theta_{\text{gene}} \, d\theta_{\text{cell}} \, d\theta_{\text{tech}} \label{eq-independent-integration}
\end{align}

@@

- $\theta_{\text{gene}} = \{\gamma^*_i, R_{\text{on},i}, t^*_{on,i}, \delta^*_i, U_{0i}\}_{i=1}^G$ represents gene-specific parameters
- $\theta_{\text{cell}} = \{t^*_j\}_{j=1}^N$ represents cell-specific temporal parameters
- $\theta_{\text{tech}} = \{\lambda_j\}_{j=1}^N$ represents technical efficiency parameters

@@

1. **Sample process duration**:
   \begin{align}
   {T_M}^{(s)} &\sim \text{Gamma}(\alpha_{T}, \beta_{T})
   \end{align}

2. **Sample cell-specific temporal positions**:
   \begin{align}
   {t^{*}_j}^{(s)} &\sim \text{Uniform}(0, {T_M}^{(s)}) \quad \text{for } j \in \{1, \ldots, N\}
   \end{align}

3. **Sample gene-specific parameters**:
   \begin{align}
   {\gamma^{*}_i}^{(s)} &\sim \text{LogNormal}(\log(1.0), 0.5^2) \\
   {R_{\text{on},i}}^{(s)} &\sim \text{LogNormal}(\log(2.5), 0.4^2) \\
   {t^*_{on,i}}^{(s)} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \\
   {\delta^*_i}^{(s)} &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \\
   {U_{0i}}^{(s)} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)
   \end{align}

@@

5. **Compute latent concentrations**:
   \begin{align}
   {u^{*}_{ij}}^{(s)}, {s^{*}_{ij}}^{(s)} &= f_{\text{piecewise}}({t^*_j}^{(s)}, {\gamma^*_i}^{(s)}, {R_{\text{on},i}}^{(s)}, {t^*_{on,i}}^{(s)}, {\delta^*_i}^{(s)})
   \end{align}

@@

1. **Process duration uncertainty**: Uncertainty in $T_M$ affects the temporal scope of the biological process and constrains cell temporal positions

@@

4. **Gene-specific temporal pattern uncertainty**: Prior uncertainty in $t^*_{on,i}, \delta^*_i$ affects when activation occurs and how long it lasts in absolute time

@@

1. **Parameter uncertainty**: $\text{Var}[\theta | y_{obs}]$ from the posterior distribution over all model parameters
2. **Temporal uncertainty**: Uncertainty in latent time coordinates $t^*_j$ and process duration $T_M$
3. **Kinetic uncertainty**: Uncertainty in gene-specific kinetic parameters $\gamma^*_i$ and fold-changes $R_{\text{on},i}$
4. **Concentration uncertainty**: Uncertainty in predicted latent concentrations $u^*_{ij}(\theta), s^*_{ij}(\theta)$ arising from parameter uncertainty
5. **Observation uncertainty**: Poisson noise in final count observations

@@

- **Non-temporal gene parameters**: $\gamma^*_i \sim \text{LogNormal}(\log(1.0), 0.5^2)$, $R_{\text{on},i} \sim \text{LogNormal}(\log(2.5), 0.4^2)$
- **Independent temporal gene parameters**: $t^*_{on,i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$, $\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$
- **Characteristic scales**: $U_{0i} \sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)$
- **Process duration**: $T_M \sim \text{Gamma}(\alpha_{T}, \beta_{T})$
- **Cell-specific times**: $t^*_j \sim \text{Uniform}(0, T_M)$
- **Capture efficiencies**: $\lambda_j \sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)$

@@

$$u^*_{ij}(\theta), s^*_{ij}(\theta) = f_{\text{piecewise}}(t^*_j, \gamma^*_i, R_{\text{on},i}, t^*_{on,i}, \delta^*_i)$$

@@

1. Compute phase assignments for each cell-gene pair using ${{t^{\ast}_{j}}}^{(s)}$, ${{t^{\ast}_{on,i}}}^{(s)}$, ${{\delta^{\ast}_i}}^{(s)}$
2. Evaluate analytical solutions: ${{u^*_{ij}}}^{(s)}, {{s^*_{ij}}}^{(s)} = f_{\text{piecewise}}({{t^*_{j}}}^{(s)}, {{\gamma^*_i}}^{(s)}, {{R_{\text{on},i}}}^{(s)}, {{t^*_{on,i}}}^{(s)}, {{\delta^*_i}}^{(s)})$
3. Sample observations:

@@

1. **Kinetic parameter uncertainty**: Posterior distributions over $\gamma^*_i$ and $R_{\text{on},i}$ affect the shape and magnitude of concentration trajectories
2. **Temporal parameter uncertainty**: Uncertainty in $t^*_{on,i}$ and $\delta^*_i$ affects when activation occurs and how long it lasts in absolute time
3. **Process duration uncertainty**: Uncertainty in $T_M$ affects the temporal scope and constrains cell temporal positions
4. **Cell-specific uncertainty**: Uncertainty in individual $t^*_j$ and $\lambda_j$ affects cell-specific predictions
5. **Scale uncertainty**: Uncertainty in $U_{0i}$ affects the absolute magnitude of predicted counts

@@

1. **Recover independent piecewise parameters**: $\gamma^*_i$, $R_{\text{on},i}$, $t^*_{on,i}$, $\delta^*_i$
2. **Maintain process duration inference**: Recover $T_M$ and cell temporal positions $t^*_j$
3. **Handle realistic noise levels**: Maintain accuracy with Poisson observation noise
4. **Scale efficiently**: Demonstrate computational advantages of eliminated scaling symmetry
5. **Validate fold-change interpretation**: Ensure $R_{\text{on},i}$ estimates match true activation strength
6. **Assess dimensional consistency**: Verify elimination of parameter redundancy through fixed $\alpha^*_{\text{off}i} = 1.0$
7. **Validate parameter identifiability**: Confirm elimination of scaling symmetry through independent parameterization
8. **Assess parameter separation**: Verify clean separation between kinetic ($\gamma^*_i$, $R_{\text{on},i}$) and temporal ($t^*_{on,i}$, $\delta^*_i$) parameter inference
9. **Validate independent parameter framework**: Confirm that independent temporal parameters enable reliable parameter recovery across different experimental contexts
